---
title: "Class 8: Breast Cancer Mini-Project"
author: "Dariana Becerra Guzman (A17506182)"
format: pdf
toc: TRUE
---

## Background

In today's class we will apply the methods and techniques clustering and PCA to help make sense of a real world breast cancer FNA biopsy data set. 

## Data Import

We start by importing our data. It is a CSV file s0 we will use the `read.csv()` function. 


```{r}
# Save your input data file into your Project directory
fna.data <- "WisconsinCancer.csv"

# Complete the following code to input the data and store as wisc.df
wisc.df <- read.csv("WisconsinCancer.csv", row.names = 1)
head(wisc.df)
```


Make sure to remove the first `diagnosis` column - I dont want to use this for me machine learning models. We will use it later to compare our results to the expert diagnosis. 

```{r}
wisc.data <- wisc.df[,-1]
diagnosis <- wisc.df$diagnosis
```

## Exploratory data analysis

>Q1. How many observations are in this dataset?

```{r}
nrow(wisc.data)
```

>Q2. How many of the observations have a malignant diagnosis?

```{r}
sum(wisc.df$diagnosis == "M")
```

```{r}
table(wisc.df$diagnosis)
```


>Q3. How many variables/features in the data are suffixed with `_mean`?

```{r}
#colnames(wisc.data)

length(grep("_mean", colnames(wisc.df), value = T))
```

## Principal Component Analysis

The main function here is `prcomp()` and we want to make sure we set the optional argument `scale=TRUE`:

```{r}
wisc.pr <- prcomp(wisc.data, scale=T)
summary(wisc.pr)
```

>Q4. From your results, what proportion of the original variance is captured by the first principal component (PC1)?

0.4427
>Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

PC3
>Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

PC7

Our main PCA "score plot" or "PC plot" of results:

```{r}
library(ggplot2)
```
```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC2, col=diagnosis) +
  geom_point()
```
 
```{r}
biplot(wisc.pr)
```

>Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

It is crowded and difficult to understand. 

>Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

The plots indicate separation between the malignant and begnin samples. 

```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC3, col=diagnosis) +
  geom_point()
```

Collectively these two plots ("score plot" and "loadings plot") tells us that if cell nucleus are deeply indented ("concave"), irregular non-circular ("compactness"), and have large "perimeter" values they tend to be malignant...

>Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?

```{r}
wisc.pr$rotation["concave.points_mean", 1]
```

## Hierarchial Clustering

First scale the data `scale()`, then calculate the distance matrix `dist()`. Then cluster with the `hclust()` function and plot:

```{r}
wisc.hclust <- hclust(dist (scale(wisc.data)))
```

>Q10. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

```{r}
plot(wisc.hclust)
abline(h=20, col="red")
```
You can claso use `cutree()` function with an argument `k=4` rather than `h=height`. 

```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k=4)
table(wisc.hclust.clusters)
```

## Combining methods 

Here we will take our PCA results and use those as input for clustering. In other words, our `wisc.pr$x` scored that we plotted above (the main output from PCA- how the data lies on our new principal component axis/variables) and use a subset of the PCs that capture the most variance as input for `hclust()`. 

```{r}
pc.dist <- dist(wisc.pr$x[,1:3])
wisc.pr.hclust <- hclust(pc.dist, method = "ward.D2")
plot(wisc.pr.hclust)
```
Cut the dendrogram/tree into two main groups/clusters:

```{r}
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
```
I want to know how the clustering in `grps` with values of 1 or 2 corespond to the expert `diagnosis`.

```{r}
table(grps,diagnosis)
```
My clustering **groups 1** are mostly "M" diagnosis (179), and my clustering **group 2** are mostly "B" diagnosis (333).

>Q15. OPTIONAL: Which of your analysis procedures resulted in a clustering model with the best specificity? How about sensitivity?

- 24 FP
- 179 TP
- 333 TN
- 33 FN

Sensitivity TP/(TP+FN)
```{r}
179/(179+33)
```

Specificity TN/(TN+FP)
```{r}
333/(333+24)
```


```{r}
ggplot(wisc.pr$x) +
  aes(PC1, PC2) +
  geom_point(col=grps)
```
## Prediction

```{r}
#url <- "new_samples.csv"
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```

```{r}
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```
>Q16. Which of these new patients should we prioritize for follow up based on your results?

Patients in the 1 category. 




