---
title: "class05: data viz with ggplot"
author: "Dariana (PID: A17506182)"
format: pdf
---

## Background

There are lot's of ways to make figures in R. These include so-called "base-R" graphics (e.g. `plot()`) and tones of add-on packages like **ggplot2**.

For example here we make the same plot with both:

```{r}
head(cars)
```

```{r}
plot(cars)
```
First I need to install the package with the command
`install.packages()`. 

> **N.B.** We never run an install cmd in a quarto code chunk or we will end up re-installing packages many many times - which is not what we want! 

Everytime we want to use one of these "add-on" packages we need to load it up in the R with the `library()` function: 

```{r}
library(ggplot2) 
```
```{r}
ggplot(cars)
```

Every ggplot needs at least 3 things: 

- The **data**, the stuff you want plotted
- The **aes**thitics, how the data map to the plot 
- The **geom**etry, the type of plot

```{r}
ggplot(cars) + 
  aes(x=speed, y=dist) +
  geom_point() 
```
Add a line to better show relationship between speed and dist
```{r}
p <- ggplot(cars) + 
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth(method="lm", se=FALSE) +
  labs(title="Stopping Distance of Old Cars", subtitle= "Data from the `cars` object", x="Speed (mph)", y="Distance (ft)")
```

```{r}
p + theme_bw()
```


## Gene Expression Plot

We can read the input data from the class website

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

A first version plot

```{r}
ggplot(genes) +
  aes(Condition1, Condition2) +
  geom_point()
```
```{r}
table(genes$State)
```

Version 2 let's color by `State` so we can see the "up" and "down" significant genes compared to all the unchanging genes

```{r}
ggplot(genes) +
  aes(Condition1, Condition2, col=State) +
  geom_point()
```

Version 3 plot, let's modify the default colors to something we like


```{r}
ggplot(genes) +
  aes(Condition1, Condition2, col=State) +
  geom_point() +
  scale_color_manual(values= c("orange", "gray", "purple")) +
  labs(x="Control (no drug)", y="Drug", title="Gene Expression Changes Upon GLP-1 Drug") +
  theme_bw()
```


## Going Further

```{r}
# File location online
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)
```
```{r}
head(gapminder, 3)
```
```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col=continent) +
  geom_point(alpha=0.3)
```

Let's "facet" (i.e. make a seperate plot) by continent rather than the big hot mess above. 

```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col=continent) +
  geom_point(alpha=0.3) +
  facet_wrap(~continent)
```

 
## Custom Plot

How big is this gapminder dataset? 

```{r}
nrow(gapminder)
```
I want to "filter" down to a subset of this data. I will use the **dplyr** package to help me. 

First I need to install it and then load it up. 

```{r}
library(dplyr)
gapminder_2007 <- filter(gapminder, year==2007)
head(gapminder_2007)
```
```{r}
filter(gapminder, year==2007, country=="Ireland")
```
 
```{r}
filter(gapminder, year==2007, country=="United States")
```

> Q. Make a plot comparing 1997 and 2007 for all countries 

```{r}
input <- filter(gapminder, year %in% c(1997, 2007))
head(input)
```

```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col=continent)+
  geom_point()+
  filter(gapminder, year %in% c(1997, 2007)) +
  facet_wrap(~year)
```













